Using high-resolution variant frequencies to empower clinical genome interpretation

### 2016-10-07 manuscript/manuscript.Rmd
More reformatting

Corespondance should be addressed to:
Dr James Ware, Cardiovascular Genetics and Genomics, NHLI, RBH Campus, Imperial College London, London, SW3 6NP; Tel: +44 20 7352 8121; Fax: +44 20 7351 8146; Email: j.ware@imperial.ac.uk

@@

**Results**: Using the example of cardiomyopathy, we show that our approach reduces by two-thirds the number of candidate variants under consideration in the average exome, and identifies 43 variants previously reported as pathogenic that can now be reclassified.

@@

Here we present a statistical framework for assessing whether rare variants are sufficiently rare to cause penetrant Mendelian disease, while accounting for both architecture and sampling variance in observed allele counts. We demonstrate that allele frequency cutoffs well below 0.1% are justified for a variety of human disease phenotypes and that such filters can remove an additional two-thirds of variants from consideration when compared to traditionally lenient frequency cutoffs. We present pre-computed allele frequency filtering values for all variants in the Exome Aggregation Consortium database, which are now available through the ExAC data browser and for download, to assist others in applying our framework.

@@

### Calculating maximum tolerated allele counts

The maximum alelle frequency (AF) of a dominant disease-causing variant in the population was defined as:  
*maximum credible population AF = prevalence* $\times$ *maximum allelic contribution* $\times$ *1/penetrance*

Determination of the maximum allelic contribution (a measure of heterogeneity) is described in the text.

@@

Disease prevalence estimates were obtained from the literature and taken as the highest value reported. Variant penetrance of 0.5 was used for all analyses, as penetrance estimates for individual variants are not widely available. This corresponds to the reported penetrance of the HCM variant used to illustrate our approach `r citep("10.1161/CIRCRESAHA.109.216291")` and is the minimum found when researching other variants/disorders.

The maximum tolerated allele count (AC) was computed as the AC at the upper bound of the one-tailed 95% confidence interval (95%CI AC) of a Poisson distribution, for the established maximum credible AF, given the observed allele number (AN). In R: `max_ac = qpois(quantile_limit,an*af)`, where `max_ac` is the 95%CI AC, `quantile_limit` is 0.95 (for a one-sided 95%CI), `an` is the observed allele number, and `af` is the maximum credible population AF.

@@

322 HCM patients (confirmed by cardiac MRI) recruited to the Royal Brompton Hospital cardiac BRU were sequenced using the IlluminaTruSight Cardio Sequencing Kit `r citep("10.1007/s12265-016-9673-5")` on Illumina MiSeq and NextSeq platforms. This study had ethical approval (REC: 09/H0504/104+5) and informed consent was obtained for all subjects. The number of rare variants in _MYBPC3_, _MYH7_ and the six other sarcomeric genes associated with HCM (_TNNT2_, _TNNI3_, _MYL2_, _MYL3_, _TPM1_ and _ACTC1_) were calculated for this HCM cohort, and for reference population samples from ExAC. Variant frequencies were calculated for all protein altering variants (frameshift, nonsense, splice donor/acceptor, missense and in-frame insertions/deletions), with case/control odds ratios calculated separately for non-overlapping ExAC AF bins with the following breakpoints: 1x10^-5^, 5x10^-5^, 1x10^-4^, 5x10^-4^ and 1x10^-3^. Odds Ratios were calculated as OR = (cases with variant/cases without variant) / (ExAC samples with variant/ExAC samples without variant). In the absence of sample-level genotype data for ExAC, the number of samples with a variant was approximated by the total number of variant alleles (assuming each rare variant was from a distinct sample).

All data required to reproduce these analyses is available at <https://github.com/ImperialCardioGenetics/frequencyFilter>. The manuscript was compiled in R, and source code for the analysis, figures and manuscript, are available at the same location. Curated variant interpretations are deposited in ClinVar *Accession & DOI to be added*. ExAC annotations are available at <exac.broadinstitute.org>. Our allele frequency calculator app is located at <https://jamesware.shinyapps.io/alleleFrequencyApp/>, with source code available at <http://github.com/jamesware/alleleFrequencyApp>.

@@

*maximum credible population AF = prevalence x maximum allelic contribution x 1/penetrance*

If we turn the problem around, and begin instead from allele frequency, specifying a maximum _true_ allele frequency value we are willing to consider in the population (using the equation above), then we can estimate the probability distribution for allele counts in a given sample size. This follows a binomial distribution, and can be satisfactorily approximated with a Poisson distribution (see **Methods**). This allows us to set an upper limit on the number of alleles in a sample that is consistent with a given population frequency.

Taking a range of cardiac disorders as exemplars, we use this framework to define the maximum credible allele frequency for disease-causing variants in each condition, and define and validate a set of maximum tolerated allele counts in the ExAC reference population sample.

@@

We illustrate our general approach using the dominant cardiac disorder hypertrophic cardiomyopathy (HCM), which has an estimated prevalence of 1 in 500 in the general population `r citep("10.1038/gim.2014.205")`. As there have been previous large-scale genetic studies of HCM, with series of up to 6,179 individuals `r citep(c("10.1038/gim.2014.205","10.1038/gim.2016.90"))`, we can make the assumption that no newly identified variant will be more frequent in cases that those identified to date (at least for well-studied ancestries). This allows us to define the maximum allelic contribution of any single variant to the disorder. In these large case series, the largest proportion of cases is attributable to the missense variant _MYBPC3_ c.1504C>T (p.Arg502Trp), found in 104/6179 HCM cases (1.7%; 95%CI 1.4-2.0%) `r citep(c("10.1038/gim.2014.205","10.1038/gim.2016.90"))`. We therefore take the upper bound of this proportion (0.02) as an estimate of the maximum allelic contribution in HCM (Table 1). Our maximum expected population allele frequency for this allele, assuming 50% penetrance as previously reported `r citep("10.1161/CIRCRESAHA.109.216291")`, is 1/500 x 1/2 (dividing prevalence per individual by the number of chromosomes per individual) x 0.02 x 1/0.5 = 4.0x10^-5^, which we take as the maximum credible population AF for any causative variant for HCM (Table 1).

To apply this threshold while remaining robust to chance variation in observed allele counts, we ask how many times a variant with population allele frequency of 4.0x10^-5^ can be observed in a random population sample of a given size. For a 5% error rate we take the 95th percentile of a poisson distribution with λ = expected allele count, which is given by: sample size (chromosomes) &times; expected population allele frequency (**Methods**).  For HCM this gives us a maximum tolerated allele count of 9, assuming 50% penetrance (or 5 for fully penetrant alleles), for variants genotyped in the full ExAC cohort (sample size=121,412 chromosomes). The _MYBPC3_:c.1504C>T variant is observed 3 times in ExAC (freq=2.49x10^-5^; Table 1).

@@

@@

To assess these thresholds empirically, we explored the ExAC allele frequency spectrum of `r myVariantsN` distinct autosomal variants identified in 6179 recently published HCM cases referred for diagnostic sequencing, and individually assessed and reported according to international guidelines`r citep(c("10.1038/gim.2014.205","10.1038/gim.2016.90"))`. 
`r myPassPath`/`r myAllPath` (`r signif(100*myPassPath/myAllPath,3)`%) variants reported as ‘Pathogenic’ or ‘Likely Pathogenic’ fell below our threshold (Figure 1), including all variants with a clear excess in cases. `r myAbsentPath` of these variants are absent from ExAC.  The `r myFailPath` variants historically classified as ‘Likely Pathogenic’, but prevalent in ExAC in this analysis, were reassessed using contemporary ACMG criteria: there was no strong evidence in support of pathogenicity, and they were reclassified in light of these findings (Table S1). This analysis identifies `r myFailVus`/`r myAllVus` (`r signif(100*myFailVus/myAllVus,3)`%) VUS that are unlikely to be truly causative for HCM.

@@

@@

This approach can be readily applied in diseases where large case series are available to assess the genetic and allelic architecture, such as the inherited cardiac conditions displayed in Table 1. In the absence of large case series, we must estimate the genetic architecture parameters by extrapolating from similar disorders and/or variant databases.

@@

@@

Thirdly, while we believe that ExAC is depleted of severe childhood inherited conditions, and not enriched for cardiomyopathies, it could be enriched relative to the general population for some conditions, including Mendelian forms of common diseases such as diabetes or coronary disease that have been studied in contributing cohorts. Where this is possible, the maximum credible population allele frequency can be simply computed based on the estimated disease prevalence in the ExAC cohort, rather than the population prevalence. Finally, although the resulting allele frequency thresholds are more stringent than those previously used, they are likely to still be very lenient for many applications. For instance, we base our calculation on the most prevalent known pathogenic variant from a disease cohort. For HCM, for which more than 6,000 people have been sequenced, it is unlikely that any single newly identified variant, not previously catalogued in this large cohort, will explain a similarly large proportion of the disease as the most common causal variant, at least in well-studied populations. Future work may therefore involve modeling the frequency distribution of all known variants for a disorder, to further refine these thresholds.

>>>>

Corespondance should be addressed to:
Dr James Ware, Cardiovascular Genetics and Genomics, National Heart and Lung Institute, Royal Brompton Hospital Campus, Imperial College London, London, SW3 6NP; Tel: +44 20 7352 8121; Fax: +44 20 7351 8146; Email: j.ware@imperial.ac.uk

@@

**Results**: Using the example of cardiomyopathy, we show that our approach reduces by two-thirds the number of candidate variants under consideration in the average exome, without removing true pathogenic variants (false positive rate<0.001).

@@

Here we present a statistical framework for assessing whether rare variants are sufficiently rare to cause penetrant Mendelian disease, while accounting for both architecture and sampling variance in observed allele counts. We demonstrate that allele frequency cutoffs well below 0.1% are justified for a variety of human disease phenotypes and that such filters can remove an additional two-thirds of variants from consideration when compared to traditionally lenient frequency cutoffs, without discarding true pathogenic variants. We present pre-computed allele frequency filtering values for all variants in the Exome Aggregation Consortium database, for comparison with user-defined thresholds, which are now available through the ExAC data browser and for download, to assist others in applying our framework.

@@

*maximum credible population AF = prevalence* $\times$ *maximum allelic contribution* $\times$ *1/penetrance*

@@

Disease prevalence estimates were obtained from the literature and taken as the highest value reported. Variant penetrance of 0.5 was used for all analyses, as penetrance estimates for individual variants are not widely available. This corresponds to the reported penetrance of the HCM variant used to illustrate our approach `r citep("10.1161/CIRCRESAHA.109.216291")` and is the minimum found when researching other variants/disorders. When the population frequency of a variant is known, we can use this to estimate prevalence `r citep("10.1126/scitranslmed.aad5169")`.

If we turn the problem around, and begin instead from allele frequency, specifying a maximum _true_ allele frequency value we are willing to consider in the population (using the equation above), then we can estimate the probability distribution for allele counts in a given sample size. This follows a binomial distribution, and can be satisfactorily approximated with a Poisson distribution (see **Supplementary Methods**). This allows us to set an upper limit on the number of alleles in a sample that is consistent with a given population frequency.

The maximum tolerated allele count (AC) was therefore computed as the AC at the upper bound of the one-tailed 95% confidence interval (95%CI AC) of a Poisson distribution, for the established maximum credible AF, given the observed allele number (AN). In R: `max_ac = qpois(quantile_limit,an*af)`, where `max_ac` is the 95%CI AC, `quantile_limit` is 0.95 (for a one-sided 95%CI), `an` is the observed allele number, and `af` is the maximum credible population AF.

@@

322 HCM patients (confirmed by cardiac MRI) recruited to the NIHR Royal Brompton cardiovascular BRU were sequenced using the IlluminaTruSight Cardio Sequencing Kit `r citep("10.1007/s12265-016-9673-5")` on Illumina MiSeq and NextSeq platforms. This study had ethical approval (REC: 09/H0504/104+5) and informed consent was obtained for all subjects. The number of rare variants in _MYBPC3_, _MYH7_ and the six other sarcomeric genes associated with HCM (_TNNT2_, _TNNI3_, _MYL2_, _MYL3_, _TPM1_ and _ACTC1_) were calculated for this HCM cohort, and for reference population samples from ExAC. Variant frequencies were calculated for all protein altering variants (frameshift, nonsense, splice donor/acceptor, missense and in-frame insertions/deletions), with case/control odds ratios calculated separately for non-overlapping ExAC AF bins with the following breakpoints: 1x10^-5^, 5x10^-5^, 1x10^-4^, 5x10^-4^ and 1x10^-3^. Odds Ratios were calculated as OR = (cases with variant/cases without variant) / (ExAC samples with variant/ExAC samples without variant). In the absence of sample-level genotype data for ExAC, the number of samples with a variant was approximated by the total number of variant alleles (assuming each rare variant was from a distinct sample).

@@

@@

We illustrate our general approach using the dominant cardiac disorder hypertrophic cardiomyopathy (HCM), which has an estimated prevalence of 1 in 500 in the general population `r citep("10.1038/gim.2014.205")`. As there have been previous large-scale genetic studies of HCM, with series of up to 6,179 individuals `r citep(c("10.1038/gim.2014.205","10.1038/gim.2016.90"))`, we can make the assumption that no newly identified variant will be more frequent in cases that those identified to date (at least for well-studied ancestries). This allows us to define the maximum allelic contribution of any single variant to the disorder. In these large case series, the largest proportion of cases is attributable to the missense variant _MYBPC3_ c.1504C>T (p.Arg502Trp), found in 104/6179 HCM cases (1.7%; 95%CI 1.4-2.0%) `r citep(c("10.1038/gim.2014.205","10.1038/gim.2016.90"))`. We therefore take the upper bound of this proportion (0.02) as an estimate of the maximum allelic contribution in HCM (Table 1). Our maximum expected population allele frequency for this allele, assuming 50% penetrance as previously reported `r citep("10.1161/CIRCRESAHA.109.216291")`, is 1/500 x 1/2 (dividing prevalence per individual by the number of chromosomes per individual) x 0.02 x 1/0.5 = 4.0x10^-5^, which we take as the maximum credible population AF for any causative variant for HCM (Table 1). 
To apply this threshold while remaining robust to chance variation in observed allele counts, we ask how many times a variant with population allele frequency of 4.0x10^-5^ can be observed in a random population sample of a given size. For a 5% error rate we take the 95th percentile of a poisson distribution with λ = expected allele count, which is given by: sample size (chromosomes) &times; expected population allele frequency (**Methods**). For HCM this gives us a maximum tolerated allele count of 9, assuming 50% penetrance (or 5 for fully penetrant alleles), for variants genotyped in the full ExAC cohort (sample size=121,412 chromosomes). The _MYBPC3_:c.1504C>T variant is observed 3 times in ExAC (freq=2.49x10^-5^; Table 1).

@@

We computed the maximum tolerated allele count for four additional cardiac disorders with available cases series. In all cases the commonest variant in the cases series was well within the calculated maximum alelle count in ExAC (Table 1).

@@

To assess the HCM thresholds empirically, we explored the ExAC allele frequency spectrum of `r myVariantsN` distinct autosomal variants identified in 6179 recently published HCM cases referred for diagnostic sequencing, and individually assessed and reported according to international guidelines`r citep(c("10.1038/gim.2014.205","10.1038/gim.2016.90"))`. 
`r myPassPath`/`r myAllPath` (`r signif(100*myPassPath/myAllPath,3)`%) variants reported as ‘Pathogenic’ or ‘Likely Pathogenic’ fell below our threshold (Figure 1), including all variants with a clear excess in cases. `r myAbsentPath` of these variants are absent from ExAC.  The `r myFailPath` variants historically classified as ‘Likely Pathogenic’, but prevalent in ExAC in this analysis, were reassessed using contemporary ACMG criteria: there was no strong evidence in support of pathogenicity, and they were reclassified in light of these findings (Table S1). This analysis identifies `r myFailVus`/`r myAllVus` (`r signif(100*myFailVus/myAllVus,3)`%) VUS that are unlikely to be truly causative for HCM.

@@

After curating variants above our calculated HCM threshold, the false discovery rate was 0/477 (0.000; 95%CI 0.000-0.008) and 1/559 (0.002; 95%CI 0.000-0.010) for the published HCM cohort and ClinVar data respectively.

@@

In the absence of large case series, we must estimate the genetic architecture parameters by extrapolating from similar disorders and/or variant databases.

@@

All data required to reproduce these analyses is available at <https://github.com/ImperialCardioGenetics/frequencyFilter>. The manuscript was compiled in R, and source code for the analysis, figures and manuscript, are available at the same location. Curated variant interpretations are deposited in ClinVar *Accession & DOI to be added*. ExAC annotations are available at <https://exac.broadinstitute.org>. Our allele frequency calculator app is located at <https://jamesware.shinyapps.io/alleleFrequencyApp/>, with source code available at <http://github.com/jamesware/alleleFrequencyApp>.

@@

Thirdly, while we believe that ExAC is depleted of severe childhood inherited conditions, and not enriched for cardiomyopathies, it could be enriched relative to the general population for some conditions, including Mendelian forms of common diseases such as diabetes or coronary disease that have been studied in contributing cohorts. Where this is possible, the maximum credible population allele frequency should be derived based on the estimated disease prevalence in the ExAC cohort, rather than the population prevalence. Finally, although the resulting allele frequency thresholds are more stringent than those previously used, they are likely to still be very lenient for many applications. For instance, we base our calculation on the most prevalent known pathogenic variant from a disease cohort. For HCM, for which more than 6,000 people have been sequenced, it is unlikely that any single newly identified variant, not previously catalogued in this large cohort, will explain a similarly large proportion of the disease as the most common causal variant, at least in well-studied populations. Future work may therefore involve modeling the frequency distribution of all known variants for a disorder, to further refine these thresholds.
